---
title: "UMR delay hours vs. INRIX congestion hours"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(readxl)
library(janitor)
library(stringr)
library(modelr)
library(dplyr)
library(stringr)
library(psych)
options(scipen=999)
```

# UMR 2020 vs INRIX 2021 - RAW
UMR report has 2020 data of Annual Hours of Delay per auto commuter. This data looks quite similar to INRIX data in terms of distribution. An advantage of UMR is its fullness (INRIX that has 50% missing data). 

```{r, echo = FALSE}
#sent by Baruch
umr_2020 <- rio::import("Copy of complete-data-2021-umr-by-tti.xlsx") %>% filter(Year == "2020") %>% 
  rename(auto_commuters = Auto, # original column:  Auto Commuter (000)
    total_delay_person_hours = `Annual Hours of Delay`, # original column: " Total Delay (Person-Hours) (000) " --> Not clear what this is
    hours_delay_perAuto_commuter = `...24`, # original column: per Auto Commuter
    population = `Population...6`, 
    city = `Urban Area`, 
    state = Primary) %>% 
  
  select(city, state, auto_commuters, population, hours_delay_perAuto_commuter, total_delay_person_hours) %>% 
  mutate(auto_commuters = as.numeric(auto_commuters)*1000,
         total_delay_person_hours = as.numeric(total_delay_person_hours)*1000,
         population = as.numeric(population)*1000,
        hours_delay_perAuto_commuter = as.numeric(hours_delay_perAuto_commuter)) 

#umr_2020 %>% arrange(desc(hours_delay_perAuto_commuter))
```
View some states/ cities: UMR shows more cities in each state. 
```{r, echo=FALSE, include=FALSE}
umr_2020 %>% arrange(desc(hours_delay_perAuto_commuter)) %>% select(city, state, hours_delay_perAuto_commuter) %>% 
  filter(state == "CA") 
```

```{r, include=FALSE}
inrix <- read_csv("inrix data 2021.csv")
inrix %>% arrange(desc(hours_lost_in_congestion)) %>% filter(state == "CA")
```

# Processed UMR data vs INRIX data

Note 1: the following factors is readily recorded in UMR >< diff from congestion_data_summary - which uses adjusted calculation --> maybe more chance of measurements error?
Note 2: ACS commuter data here is for year 2019. Why not use the commuter data in UMR 2020? 
Also, no need to use "HM-74 Daily Vehicle Miles Travelled" because UMR already has total commuters

```{r, echo=TRUE}
summary_umr_2020 <- umr_2020 %>% 
  group_by(state) %>% 
 # mutate(state_avg_delay_hours = mean(hours_delay_perAuto_commuter))
  summarise(state_avg_delay_hours = round(mean(hours_delay_perAuto_commuter),2),
            state_tot_commuters = sum(auto_commuters), 
            state_tot_delay_hours = state_tot_commuters * state_avg_delay_hours) %>% 
  ungroup() 

summary_umr_2020 %>% head()
```
## processed INRIX data

```{r}
# TB's result: 
congestion_data_summary_TB <- readRDS("congestion_data_summary.RDS")
```
## comparison
```{r, fig.width=6, echo = TRUE}
comparison_umr_inrix <- left_join(summary_umr_2020, congestion_data_summary_TB, by = ("state")) %>% 
  rename(state_tot_commuters_umr = state_tot_commuters.x,
         state_tot_commuters_inrix = state_tot_commuters.y)

comparison_umr_inrix %>% select(state, state_avg_delay_hours, state_avg_congestion_hours) %>% 
  pivot_longer(cols = 2:3,
      names_to = "delay_congestion", 
               values_to = "value") %>% 
ggplot(aes(state, value, group = delay_congestion)) +
  geom_col(position = "dodge", aes(fill = delay_congestion)) +
  coord_flip() +
  theme_minimal()
```


```{r, echo = TRUE}
# check extreme case: NJ
comparison_umr_inrix %>% filter(state == "NJ")

# check original data
inrix %>% filter(state == "NJ") %>% mutate(avg = mean(hours_lost_in_congestion)) 
umr_2020 %>% filter(state == "NJ") %>% mutate(avg = mean(hours_delay_perAuto_commuter)) 
# mean of state_avg_congestion_hours should not be too big

# check the final result
AHR_data <- readRDS("AHR_data_INRIXdata.RDS")
AHR_data %>% filter(state == "New Jersey")
```

```{r, echo = TRUE}

# check "normal" case: WA
comparison_umr_inrix %>% filter(state == "WA")

# check original data --> the original avg should be very close to each other. 
# state_avg_congestion_hours seem to be inflated too big. 
inrix %>% filter(state == "WA") %>% mutate(avg = mean(hours_lost_in_congestion)) 
umr_2020 %>% filter(state == "WA") %>% mutate(avg = mean(hours_delay_perAuto_commuter)) 
```


# Other factors

```{r, include=FALSE}
#HM-10: Public road length 2020, miles by ownership
HM_10 <- read_excel("hm10.xls", sheet = "A") %>% 
  remove_empty() %>% 
  select(1, 2, 5, 8, 11) %>% 
  slice(-(1:6)) %>% 
  rename(state = 1,
         rural_SHA = 2,
         rural_other = 3,
         urban_SHA = 4,
         urban_other = 5) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(rural_SHA:urban_other, as.numeric),
         SHA_miles = rural_SHA + rural_other + urban_SHA + urban_other) %>% 
  select(state, SHA_miles)


#HM-81: State highway agency-owned public roads 2020; rural and urban miles, estimated lane-miles and daily travel
HM_81 <- read_excel("hm81.xls", sheet  = "A") %>% 
  remove_empty() %>% 
  select(1, 17) %>% 
  rename(state = 1,
         state_controlled_lane_miles = 2) %>% 
  slice(-(1:7)) %>% 
  filter(state %in% state.name) %>% 
  mutate(state_controlled_lane_miles = as.numeric(state_controlled_lane_miles))


#SF-4: Disbursements for state-administered highways (thousands of dollars)
#Note: the 2020 numbers for some reason exactly match the 2019 numbers? 
SF_4 <- read_excel("sf4.xlsx", sheet = "A") %>% 
  remove_empty() %>% 
  select(1:4, 8) %>% 
  rename(state = 1,
         capital_disbursement = 2,
         maintenance_disbursement = 3,
         admin_disbursement = 4,
         total_disbursement = 5) %>% 
  slice(-(1:6)) %>% 
  mutate(state = str_replace_all(state, "[:punct:]", ""),    #remove all special characters
         state = str_replace_all(state, "[:digit:]", ""),    #remove all numbers
         state = str_trim(state),
         across(capital_disbursement:total_disbursement, as.numeric)) %>% 
  filter(state %in% state.name)


#HM-64: Miles by measured pavement roughness
HM64_rural_interstate <- read_excel("hm64.xls", sheet = "A") %>% 
  remove_empty() %>% 
  select(1, 8:11) %>% 
  rename(state = 1,
         rural_interstate_171_194 = 2,
         rural_interstate_195_220 = 3,
         rural_interstate_above_220 = 4,
         rural_interstate_total = 5) %>% 
  slice(-(1:7)) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(rural_interstate_171_194:rural_interstate_total, as.numeric),
         rural_interstate_above_170 = rural_interstate_171_194 + rural_interstate_195_220 + rural_interstate_above_220)
  

HM64_rural_OPA <- read_excel("hm64.xls", sheet = "B") %>% 
  remove_empty() %>% 
  select(1, 20:21) %>% 
  rename(state = 1,
         rural_OPA_above_220 = 2,
         rural_OPA_total = 3) %>% 
  slice(-(1:7)) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(rural_OPA_above_220:rural_OPA_total, as.numeric))
  
  
HM64_urban_interstate <- read_excel("hm64.xls", sheet = "C") %>% 
  remove_empty() %>% 
  select(1, 8:11) %>% 
  rename(state = 1,
         urban_interstate_171_194 = 2,
         urban_interstate_195_220 = 3, 
         urban_interstate_above_220 = 4,
         urban_interstate_total = 5) %>% 
  slice(-(1:7)) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(urban_interstate_171_194:urban_interstate_total, as.numeric),
         urban_interstate_above_170 = urban_interstate_171_194 + urban_interstate_195_220 + urban_interstate_above_220)


HM64_urban_OPA <- read_excel("hm64.xls", sheet = "D") %>% 
  remove_empty() %>% 
  select(1, 20:21) %>% 
  rename(state = 1,
         urban_OPA_above_220 = 2,
         urban_OPA_total = 3) %>% 
  slice(-(1:7)) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(urban_OPA_above_220:urban_OPA_total, as.numeric))
  
HM64 <- HM64_rural_interstate %>% 
  left_join(HM64_rural_OPA) %>% 
  left_join(HM64_urban_interstate) %>% 
  left_join(HM64_urban_OPA) %>% 
  select(state,
         rural_interstate_above_170,
         rural_interstate_total,
         rural_OPA_above_220,
         rural_OPA_total,
         urban_interstate_above_170,
         urban_interstate_total,
         urban_OPA_above_220,
         urban_OPA_total)


#FI-20: Persons fatally injured in motor vehicle crashes
FI_20 <- read_excel("fi20.xls", sheet = "A") %>% 
  remove_empty() %>% 
  select(1:4, 10:12, 19) %>% 
  rename(state = 1,
         rural_interstate_fatalities = 2,
         rural_OFE_fatalities = 3,
         rural_OPA_fatalities = 4,
         urban_interstate_fatalities = 5,
         urban_OFE_fatalities = 6,
         urban_OPA_fatalities = 7,
         total_fatalities = 8) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(rural_interstate_fatalities:total_fatalities, as.numeric),
         rural_I_OFE_OPA_fatalities = rural_interstate_fatalities + rural_OFE_fatalities + rural_OPA_fatalities,
         urban_I_OFE_OPA_fatalities = urban_interstate_fatalities + urban_OFE_fatalities + urban_OPA_fatalities) %>% 
  select(state, total_fatalities, rural_I_OFE_OPA_fatalities, urban_I_OFE_OPA_fatalities)


#VM-2: Annual vehicle miles
VM_2 <- read_excel("vm2.xls", sheet = "A") %>%
  select(1:4, 10:12, 18) %>% 
  rename(state = 1,
         rural_interstate_VMT = 2,
         rural_OFE_VMT = 3,
         rural_OPA_VMT = 4,
         urban_interstate_VMT = 5,
         urban_OFE_VMT = 6,
         urban_OPA_VMT = 7,
         total_VMT = 8) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(rural_interstate_VMT:total_VMT, as.numeric),
         rural_VMT = rural_interstate_VMT + rural_OFE_VMT + rural_OPA_VMT,
         urban_VMT = urban_interstate_VMT + urban_OFE_VMT + urban_OPA_VMT) %>% 
  select(state, total_VMT, rural_VMT, urban_VMT)

#Bridge data
bridge_raw <- read_excel("fccount21.xlsx", sheet = "2021")  

bridge_total <- bridge_raw[1:58,] %>% 
  rename(state = 1) %>% 
  mutate(state = str_to_title(state)) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(2:13, as.numeric)) %>% 
  mutate(total_bridges = rowSums(across(where(is.numeric)))) %>% 
  select(state, total_bridges)

bridge_poor <- bridge_raw[180:nrow(bridge_raw),] %>% 
  rename(state = 1) %>% 
  mutate(state = str_to_title(state)) %>% 
  filter(state %in% state.name) %>% 
  mutate(across(2:13, as.numeric)) %>% 
  mutate(total_poor_bridges = rowSums(across(where(is.numeric)))) %>% 
  select(state, total_poor_bridges)

bridge_data <- bridge_total %>% 
  left_join(bridge_poor)
```

# Calculate national metrics

```{r, include=FALSE}
state_names <- data.frame(state.abb, state.name)     #need this to get state full names for congestion data

  delay_summary_umr_2020 <- summary_umr_2020 %>% 
  left_join(state_names, by = c("state" = "state.abb")) %>% 
  select(-state) %>% 
  rename(state = state.name) %>% 
  filter(!is.na(state))

#Bring everything together
AHR_data <- list(HM_10, 
                 HM_81, 
                 SF_4, 
                 HM64, 
                 FI_20, 
                 VM_2, 
                 bridge_data, 
                 delay_summary_umr_2020) %>% 
  reduce(left_join, by = "state")

#Calculate national metrics
AHR_data_national <- AHR_data %>%
  summarise(across(2:ncol(AHR_data), sum)) %>% 
  mutate(state = "United States")

AHR <- bind_rows(AHR_data, AHR_data_national) %>% 
#Add SHA ratio
  mutate(SHA_ratio = state_controlled_lane_miles / SHA_miles, .after = state_controlled_lane_miles)

```

```{r}
#Calculate key metrics
AHR_data <- AHR %>% 
  mutate(
    capital_disbursement_per_lm = capital_disbursement / state_controlled_lane_miles * 1000,
    
    maintenance_disbursement_per_lm = maintenance_disbursement / state_controlled_lane_miles * 1000,
    
    admin_disbursement_per_lm = admin_disbursement / state_controlled_lane_miles * 1000,
    
    total_disbursement_per_lm = total_disbursement / state_controlled_lane_miles * 1000,
    
    rural_interstate_poor_percent = rural_interstate_above_170 / rural_interstate_total * 100,
    
    urban_interstate_poor_percent = urban_interstate_above_170 / urban_interstate_total * 100,
    
    rural_OPA_poor_percent = rural_OPA_above_220 / rural_OPA_total * 100,
    
    urban_OPA_poor_percent = urban_OPA_above_220 / urban_OPA_total * 100,
    
    poor_bridges_percent = total_poor_bridges / total_bridges * 100,
    
    total_fatalities_per_100m_VMT = total_fatalities / total_VMT * 100,
    
    rural_fatalities_per_100m_VMT = rural_I_OFE_OPA_fatalities / rural_VMT * 100,
    
    urban_fatalities_per_100m_VMT = urban_I_OFE_OPA_fatalities / urban_VMT * 100
    
   # state_avg_delay_hours = state_tot_delay_hours/state_tot_commuters already have in umr
  )


#Calculate overall scores
AHR_data_composite <- AHR_data %>% 
  pivot_longer(cols = capital_disbursement_per_lm:state_avg_delay_hours, 
               names_to = "key_metrics", 
               values_to = "value") %>% 
  arrange(key_metrics) %>% 
  select(state, key_metrics, value) %>% 
  group_by(key_metrics) %>% 
  mutate(relative_score = value / value[state == "United States"]) %>% 
  ungroup() %>% 
  group_by(state) %>% 
  summarise(overall_score = mean(relative_score, na.rm = T)) %>% 
  ungroup()

#Add overall scores to AHR_data
AHR_data <- AHR_data %>% 
  left_join(AHR_data_composite)

#Rank states according to the calculated metrics
AHR_data_rank <- AHR_data %>% 
  filter(state != "United States") %>% 
  mutate(
    across(capital_disbursement_per_lm:overall_score, min_rank, .names = "{.col}_rank")
  )

AHR_2020 <- AHR_data %>% 
  left_join(AHR_data_rank)

#saveRDS(AHR_2020, "AHR_2020.RDS")
#write.csv(AHR_2020, "AHR_AHR_2020_UMRdata.csv")
```


